Scaling up a GF grammar for free parsing

### 2011-12-08 doc/report/report.tex
report

% check references! lncs is good maybe. lnai
\newpage
\chapter{Introduction}
\section{Introduction} 
>>This work is part of bigger project, spring and next spring.
>>Connection between grammar and parser: GF, every grammar defines a parser.
>> reusablilty, freely available.
>> the big amount of data, need language technology, machine translation
>> parsing: assingn of morph. and syntactic structures.
>>Connection between grammar and parser: GF, every grammar defines a parser.
Ultimate goal: want computers to 'understand' language. Not good with flat strings.
We need to translate strings of natural language to a richer structure,
a parse tree. So far there is no
freely available ruled-based parsing giving a deep analyse for Swedish.
%There are parsers which put only put part-of-speach tags to
%the text, and others that gives a deeper analyse. 
There are statistical parser, and parser which only are deal with controlled language.
The GF parser are one of those. 
%What is parsing, flat strings to trees, example picture, what is deep parsing.
So far stable and good when working with controlled languages, but recent experiments
have showed that it is also possible to parse free language.
To do this for Swedish, we need to  .. the resource grammar needed to covering
the language specific constructions. Further a large lexicon is also needed, as well
as a method for evaluation.
During this project, we have worked on and explored those things and some problems identified.
We have extended the Swedish resource grammar in GF, to give a better coverage and imported
a large lexicon. Work to compare the trees from Talbanken to GF and to translate them.

@@

Other work exist, but none is freely available. We want to create a open source
product, therefore can only use resources that allows this. Work can continue,
will not disappear, be made use of in other projects.
%What we could not cover by this project, what have been covered before (in
%the spring project), and why we choose those parts.
To build a parser or a grammar from scratch is a heavy work and we
therefore combine already existing parts.

@@

\subsection{Grammatical Framework}
The main component of the the project is the Grammatical Framework\cite{gfbok} (GF). It is
a grammar formalism based on Martin-Löf type theory\cite{martinlof}. GF is
designed for multilingual applications and is equivalent to Parallel Multiple
Context-Free Grammars (Seki \& al. 1991), it is hence stronger than mildly
contex-free grammars.

GF is a functional programming languages, inspired by programming languages
like ML\cite{ml}, Haskell\cite{haskell}, Prolog\cite{bratko} and
Agda\cite{agda}. 
It has been developed for about 20 years 
>> snd: functional and programming language, type checking and logic: prolog (check this)
>> dependent and implicit types : agda. (read gf book and krasses book)

A grammar written in GF can be used for both parsing and generation.
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system. The interpreter, the GF shell, lets
you try and test your grammar by using commands for parsing, vizualization of parse trees,
random generation, word alignment, morphological quizzes etc.
The compiler lets you compile your grammar into portable grammar format (PGF)\ref{pgf}.
There are libraries for using PGF files with Java, Java script and Haskell. The grammars
may also be used in Android applications.
and are inspired by.\\

@@

\vspace{5mm}
As an open source-project, GF is constantly being developed and improved. New
languages are added, the compiler is being improved, ways of using it in more 
effecient and easier ways are added
and the possibilities to use GF in different environments
increased. There are  research about how to make better use of the dependent
types, for using ontologies\cite{ramona}(>>check!!) or generating natural
language via Montague semantics\cite{montague}.
Further, experiments on using GF for free parsing has been conducted
\cite[\textsection 4.1]{gfMech}.
This is also a future goal for our project and one of the subgoal has been to
model an important part of Swedish in GF. The starting point was the resource
grammar and the result should still have a strong connection to this.

@@

\section{Related work}
>> Much more here, important!! comparisons.
Computational ling is active area for Swedish, there are many other Swedish parser,
but grammars?
Many of the parsers not ruled based, the most known one Malt\cite{malt} ... .
However has been two, about CLE\cite{cle} and what happened to it. Logic and translation
to Swedish, S-CLE (Gambäck,1997) full analyse and semantics, translation,
but coverage is too limited (nivres papper), The
spoken lang. translator\cite{cle2}. Wiren \cite{wiren}?? chartparser, but for swedish. bit more information than
just pos: cassSwe \cite{casswe}
Other Sw. parsers and grammars, evaluation of them? 
Other grammar frameworks.
SweCG (birn) annotation, parser Swedish FDG (voultanien,2001) which is not freely available.
finding subject, objects, suubject complements (SC) and
to link them to their proper regents (main verbs) : over 90\% both precision and recall.
uses a dependency grammar, but "The functional description of adverb phrases
and prepositional phrases (e.g. agent, source, goal, benefactive, time) remains
to be described in a future version." 
HPSG\cite{hpsg} and lingo matrix\cite{lingomatrix}, English, Japanese, German,
Greece, Spanish, Norweigan (NorSource, Hellan 2003),( not famous: SweCore
(Ahrenberg) based on NorSource)
also Swedish: BiTSE\cite{stymne}, Stymne 2006, Swedish English translation,
and one for the common parts of scandinavian (Haugereid etc) in Nordisk sprog.
focusing on verb phrames, built on HPSG.
S\o rgaard and Haugereid, Scandinavian Grammar Matrix\cite{scandmatrix}
Tag, Tree adjoining grammar, XTAG: English and Korean, Chinese? English: ongoing to get widecoverage grammar
under GPL. Translation to Korean.
% related and references work: CLE, Krasimirs book, CassSwe, GF book,SAG, gunlög,
% the multilingual treebank, Homse&Hinchcliff, Aarne i LILT

@@

The kind of noun phrases that can be used seems to coincide with the ones
that can be used with formal subjects (see section `Impersonal constructions').

% agreement to subj: Also good for double functions : Flickan, hon var inte dum, hon?
% F1,2,3 : NP -> VP -> Cl ; NP needs gender (Masc,Fem,Utr,Neutr)

@@

\section{Conclusion}
>> why is this significant, other grammars can use this Swedish grammar.
>> well known (comp. ling) trade-off between depth and coverage. for now depth, robustness in in future
>> work leads to coverage.
%There is obviously a lot do still, will never get finish with NPL.
%Grammar may need even bigger changes, to allow enough but not too much.
%Different from writing a grammar for generation.
%Results from mapping shows that the translation is doable, but becomes harder
%since talbanken don't have formal rules.
%The saldo shows how easy to make use of utomstående resources, promising if
%we want to extract other information. 
%Hard, but not finished!

>>>>

% check references! lncs is good maybe. lnai
\newpage
\chapter{Introduction}
%\section{Introduction} 
The grammar formalism GF has so far been succesfull used for 
controlled languges, but recent experiments have showed
that it is also possible to use the framework for parsing free language.
Parsing, that is the task of automatically identifiyng morpohlogical and
syntactical structures to text, is subject to increasing interest, especially
considering the steadily growing amount of data available online. 
So far there is no freely available ruled-based parser that gives a deep analyse for
Swedish.  There are other parsers, but the fact that they are not freely available does not only
restrict the usage but also their possiblities of being developed. Our goal is
to create a open source Swedish grammar and parser that can be continuosly enhanced and
be made use of in other projects.
To build a parser or a grammar from scratch is a heavy work and we
therefore combine already existing parts.
We utgår från GF resource grammar and therefore have 
The usage of GF allows us to start from
a well-defined system for describing grammar, as well as a strong connection to
more than 20 other langugages implentented in the framework.
The grammar defines a parser, which accepts all constructions defined by the grammar.

We hence needed to extend the Swedish grammar to cover
the language specific constructions. Further a large lexicon needed, as well
as a method for evaluation.
During this project, we have worked on and explored those things and some problems identified.
%We have extended the Swedish resource grammar in GF, to give a better coverage and imported
%a large lexicon. Work to compare the trees from Talbanken to GF and to translate them.

%Ultimate goal: want computers to 'understand' language. Not good with flat strings.
%We need to translate strings of natural language to a richer structure,
%a parse tree. 
%There are parsers which put only put part-of-speach tags to
%the text, and others that gives a deeper analyse. 
%There are statistical parser, and parser which only are deal with controlled language.
%The GF parser are one of those. 
%What is parsing, flat strings to trees, example picture, what is deep parsing.

@@

%What we could not cover by this project, what have been covered before (in
%the spring project), and why we choose those parts.

@@

\subsection{Grammatical Framework}
>> extensnion to CFG, same expressiveness as TAG and HG. parse in polynomial time.
>> incremental parsing.
The main component of the the project is the Grammatical Framework\cite{gfbok} (GF). It is
a grammar formalism based on Martin-Löf type theory\cite{martinlof}. GF is
designed for multilingual applications and the formalism is equivalent to
Parallel Multiple Context-Free Grammars (Seki \& al. 1991), it is hence
stronger than mildly contex-free grammars.

GF is a functional programming languages, inspired by programming languages
like ML\cite{ml}, Haskell\cite{haskell}, Prolog\cite{bratko} and
Agda\cite{agda}. 
>> GF ~ 15 years. PGF= runtime systm
It has been developed for about 20 years 
>> snd: functional and programming language, type checking and logic: prolog (check this)
>> dependent and implicit types : agda. (read gf book and krasses book)

A grammar written in GF can be used for both parsing and generation.
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system. The interpreter, the GF shell, lets
you try and test your grammar by using commands for parsing, vizualization of parse trees,
random generation, word alignment, morphological quizzes etc.
The compiler lets you compile your grammar into portable grammar format (PGF)\ref{pgf}.
There are libraries for using PGF files with Java, Java script and Haskell. The grammars
>> Haskell only complete, also or C\# and C. 
may also be used in Android applications.
and are inspired by.\\

@@

\vspace{5mm}
As an open source-project, GF is constantly being developed and improved. New
languages are added, the compiler is being improved, ways of using it in more 
effecient and easier ways are added
and the possibilities to use GF in different environments
increased. There are  research about how to make better use of the dependent
types, for using ontologies\cite{ramona} or generating natural
language via Montague semantics\cite{montague}.
Further, experiments on using GF for free parsing has been conducted
\cite[\textsection 4.1]{gfMech}.
This is also a future goal for our project and one of the subgoal has been to
model an important part of Swedish in GF. The starting point was the resource
grammar and the result should still have a strong connection to this.

@@

\section{Related work}
>> Much more here, important!! comparisons.
>> HPSG and LFG are context sensitive
Computational ling is active area for Swedish, there are many other Swedish parser,
but grammars?
Many of the parsers not ruled based, the most known one Malt\cite{malt} ... .
However has been two, about CLE\cite{cle} and what happened to it. Logic and translation
to Swedish, S-CLE (Gambäck,1997) full analyse and semantics, translation,
but coverage is too limited (nivres papper), The
spoken lang. translator\cite{cle2}. Wiren \cite{wiren}?? chartparser, but for swedish. bit more information than
just pos: cassSwe \cite{casswe}
Other Sw. parsers and grammars, evaluation of them? 
Other grammar frameworks.
SweCG (birn) annotation, parser Swedish FDG (voultanien,2001) which is not freely available.
finding subject, objects, suubject complements (SC) and
to link them to their proper regents (main verbs) : over 90\% both precision and recall.
uses a dependency grammar, but "The functional description of adverb phrases
and prepositional phrases (e.g. agent, source, goal, benefactive, time) remains
to be described in a future version." 
HPSG\cite{hpsg} and lingo matrix\cite{lingomatrix}, English, Japanese, German,
Greece, Spanish, Norweigan (NorSource, Hellan 2003),( not famous: SweCore
(Ahrenberg) based on NorSource)
also Swedish: BiTSE\cite{stymne}, Stymne 2006, Swedish English translation,
and one for the common parts of scandinavian (Haugereid etc) in Nordisk sprog.
focusing on verb phrames, built on HPSG.
S\o rgaard and Haugereid, Scandinavian Grammar Matrix\cite{scandmatrix}
Tag, Tree adjoining grammar, XTAG: English and Korean, Chinese? English: ongoing to get widecoverage grammar
under GPL. Translation to Korean.
% related and references work: CLE, Krasimirs book, CassSwe, GF book,SAG, gunlög,
% the multilingual treebank, Homse&Hinchcliff, Aarne i LILT

@@

The kind of noun phrases that can be used seems to coincide with the ones
that can be used with formal subjects (see section `Impersonal constructions').
>> subject controlling verbs
>> why comlicated with doubled types, show simplicity (elegant) with dep.types, and
>> some problems (many fields).
>> samesame as Obj, but conjunction works too, sin is Quant but what i sig?
>> (show circle (sig : NP -> GenNP sig -> sin??)
% agreement to subj: Also good for double functions : Flickan, hon var inte dum, hon?
% F1,2,3 : NP -> VP -> Cl ; NP needs gender (Masc,Fem,Utr,Neutr)

@@

\section{Conclusion}
>> why is this significant, other grammars can use this Swedish grammar.
>> well known (comp. ling) trade-off between depth and coverage. for now depth, robustness in in future
>> work leads to coverage.
>> reusablilty, freely available.
%There is obviously a lot do still, will never get finish with NPL.
%Grammar may need even bigger changes, to allow enough but not too much.
%Different from writing a grammar for generation.
%Results from mapping shows that the translation is doable, but becomes harder
%since talbanken don't have formal rules.
%The saldo shows how easy to make use of utomstående resources, promising if
%we want to extract other information. 
%Hard, but not finished!
